Models of lens candidates from Space Warps CFHTLS

### 2017-03-16 parts/intro.tex
editing intro

Light deflection at the rim of the Sun due to its gravitational field is famously $1.75''$.  The
deflection angle can also be expressed in terms of the escape velocity
$2v_{\rm esc}^2/c^2$, with $v_{\rm esc}$ at the solar surface being
$\simeq620\rm\,km/s$.  The Galactic escape velocity at the Sun's
location is similar: $\approx500\rm\,km/s$, typical of massive
galaxies.  Thus, the outer regions of massive galaxies have lensing
deflections comparable to that at the rim of the Sun.  Yet these
quantitatively similar deflections have qualitatively different
consequences.  Light deflection by the Sun must be accounted for in
modern astrometry \citep[see e.g.,][]{2015CQGra..32p5008C} but is not
itself a physics probe in the way it was a century ago.  With distant
galaxies, however, light bending by $\sim1''$ introduces two new
phenomena.  First, for galaxies distant enough that their apparent
size is comparable to the bending angle, light deflection can cause
multiple images of background sources, or strong lensing.  Second, the
gravitational field is dominated by dark matter, hence strong lensing
becomes a probe of dark matter in galaxies.

With no unambiguous dark-matter particle detections so far, dark
matter has been studied only through its indirect consequences on galactic
and larger scales.  In most work, dark matter is taken to be a
collisionless non-relativistic fluid (cold dark matter or CDM); in
recent years, the millennium simulation by \cite{2005Natur.435..629S} has been
particularly influential in studying the formation of CDM structures.
Other scenarios have also been considered, such as a cold condensing
boson fluid \citep{2016ApJ...818...89S} 
or even dark matter particles with internal degrees of freedom 
\citep{2011MNRAS..405...77}. %(Saxton & Ferreras 2010 MNRAS 405, 77)
There is a general
consensus, however, that 
basic mechanism of galaxy formation involve
gravitational collapse, fragmentation, and mergers of dark-matter
clumps, into which gas fell, cooling through radiative processes to
form dense clouds and eventually stars.  There is, however, much
debate about the details, of which \cite{2012RAA....12..917S} provide
a summary.

Strong gravitational lensing over galaxy scales provides
a useful source of information on the
mutual dynamics of dark matter and gas in galaxies.  The topic has
been explored in numerous studies
\citep{2009ApJ...703L..51K,2011ApJ...740...97L,2012MNRAS.424..104L,
  2016MNRAS.459.3677L,2016MNRAS.456..870B} but it is desirable to
enlarge the samples from tens of lensing galaxies to thousands.  Doing
so requires both finding more lenses and also modelling their
masses.

Recent searches through the CFHTLS \citep{2012SPIE.8448E..0MC} using
arc-finders
\citep{2012ApJ...749...38M,2014A&A...567A.111M,2014ApJ...785..144G} by
machine learning \citep{2016A&A...592A..75P} and by visual inspection
by citizen-science volunteers
\citep[Space~Warps][]{2016MNRAS.455.1191M} have, between them,
discovered an average of four lenses per square degree.  So one can be
optimistic about finding many thousands of lenses in the next
generation of wide-field surveys.  The expected flood of new lens
discoveries
with surveys such as Euclid or WFIRST,
will need a similarly huge modelling effort to reconstruct
their mass distributions.  To prepare for the challenge of
massive-sample lens modelling, \cite{2015MNRAS.447.2170K} developed a
new modelling strategy, implemented as the SpaghettiLens system.  The
idea is to collaborate with experienced members of the citizen-science
community, who have already participated in lens discovery through
Space~Warps, as well as several other projects involving astronomical
data.

>>>>

By coincidence, the typical escape velocity of massive galaxies is
such that $v_{\rm esc}^2/c^2$ is comparable to the apparent sizes of a
galaxies at cosmological distances.  This coincidence is fortunate,
because it makes the lensing deflection angle (which is $2v_{\rm
  esc}^2/c^2$) of distant galaxies comparable to their size on the
sky, and as a result, strong lensing by galaxies tends to produce
images that probe the dark halos of those galaxies.  This is
important, because while there is a general consensus that basic
mechanism of galaxy formation involve gravitational collapse,
fragmentation, and mergers of dark-matter clumps, into which gas fell,
cooling through radiative processes to form dense clouds and
eventually stars, there is much debate about the details \citep[for a
  summary, see][]{2012RAA....12..917S}.  In particular, the nature of
dark matter remains mysterious: most researchers take it to be a
collisionless non-relativistic fluid (cold dark matter or CDM) readily
studied by simulations \citep[for example, the millenium simulation
  by][which has been particularly influential]{2005Natur.435..629S}.
But other scenarios have also been considered, such as a cold
condensing boson fluid \citep{2016ApJ...818...89S}, or dark matter
particles with internal degrees of freedom
\citep{2010MNRAS.405...77S}, or not matter at all but a modification
of gravity \citep{2016PhRvL.117t1101M}.

All this motivates using galaxy lenses to study the mutual dynamics of
dark matter and gas in galaxies.  Several studies in recent years have
done so
\citep{2009ApJ...703L..51K,2011ApJ...740...97L,2012MNRAS.424..104L,
  2016MNRAS.459.3677L,2016MNRAS.456..870B} but it is desirable to
enlarge the samples from tens of lensing galaxies to thousands.  Doing
so requires both finding more lenses and also modelling their masses.
Recent searches through the CFHTLS \citep{2012SPIE.8448E..0MC} using
arc-finders
\citep{2012ApJ...749...38M,2014A&A...567A.111M,2014ApJ...785..144G} by
machine learning \citep{2016A&A...592A..75P} and by visual inspection
by citizen-science volunteers in Space~Warps
\citep{2016MNRAS.455.1191M} have, between them, discovered an average
of four lenses per square degree.  So one can be optimistic about
finding many thousands of lenses in the next generation of wide-field
surveys, from the LSST in optical and the SKA in radio on the ground,
and Euclid and WFIRST in orbit.  The expected flood of new lens
discoveries will need a similarly huge modelling effort to reconstruct
their mass distributions.

To prepare for the challenge of massive-sample lens modelling,
\cite{2015MNRAS.447.2170K} developed a new modelling strategy,
implemented as the SpaghettiLens system.  The idea is to collaborate
with experienced members of the citizen-science community, who have
already participated in lens discovery through Space~Warps, as well as
several other projects involving astronomical data.
